Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-01-24 manuscript.tex
latest

As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for our purposes. In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery), with probabilities weighted by a coefficient for each type representing competitive ability. This representation of competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}. The classic lottery model also has similarities with the Wright-Fisher model in population-genetics (section BLAH).

@@

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules''). We assume adults cannot be ousted from occupied territories, and only propagules landing on unoccupied territories are included in $m_i$. Propagules disperse at random, independently of each other, and without spatial restrictions; each propagule has an equal probability of landing on any of the $U$ unnocupied territories. Thus, there is no interaction between propagules (e.g. avoidance of territories crowded with propagules). Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts propagules which go on to contest the territory they land in). In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$. For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to dispersal to occupied territories. We assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal). This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.

@@

@@

Eq. \eqref{eq:growthsumuncoupled} gives little intuition about the dynamics of density-dependent lottery competition, since it involves an expectation over the dispersal distributions $p_i$, which depend on how the $n_i$ change over time. We now evaluate this expectation using a ``mean field'' approximation.

This approximation is similar to the high-$l_i$ approximation leading to Eq. \eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values. Replacing $x_i$ with $l_i$, as in the classic lottery model, is not valid for genotypes with low propagule densities $l_i\ll 1$ because $x_i=1$ in the few territories where $i$ propagules do land, and so $i$'s growth comes entirely from territories which deviate appreciably from the mean.

%Eq. \eqref{eq:lottery} is nonsensical if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands a fraction $c_i m_i/\sum_j c_j m_j$ of the unoccupied territories $U$, no matter how large $U$ is. The source of this pathological behavior when

The key distinction is that territories with a single propagule from the focal genotype, whose behavior is critical at low densities, are handled separately. In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion''). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq. \eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$). The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Thus, Eq. \eqref{eq:master} generalizes the classic lottery model to account for arbitrary propagule densities for each genotype.

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$). Two genotypes are present, one of which has a $c$-advantage and is at low frequency. The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b). On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq. \eqref{eq:master}. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present. (a) and (b) show the low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$). Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq. \eqref{eq:master}.} 
\end{figure}

@@

However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig.~\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

>>>>

As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for our purposes. In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery; \citealt{sale_77}), with probabilities weighted by a coefficient for each type representing competitive ability. This representation of competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}. The classic lottery model also has similarities with the Wright-Fisher model in population-genetics (section BLAH).

@@

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules''). We assume adults cannot be ousted from occupied territories, and only propagules landing on unoccupied territories are included in $m_i$. Propagules disperse at random, independently of each other, and without spatial restrictions; each propagule has an equal probability of landing on any of the $U$ unnocupied territories. Thus, there is no interaction between propagules (e.g. avoidance of territories crowded with propagules). Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts propagules which go on to contest the territory they land in). In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$. For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to dispersal to occupied territories \citep{levins_71}. We assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal). This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.

@@

Eq. \eqref{eq:lottery} is nonsensical at low densities, specifically if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands a fraction $c_i l_i/\sum_j c_j l_j$ of the unoccupied territories $U$, no matter how large $U$ is.

@@

Eq. \eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$, and is thus too complicated to give intuition about the dynamics of density-dependent lottery competition. We now evaluate this expectation using a ``mean field'' approximation.

Our approximation is similar to the high-$l_i$ approximation behind the classic lottery model in that we replace the $x_i$ with appropriate mean values. However, we cannot simply replace $x_i$ with $l_i$ as in the classic lottery model. For a genotype with a low propagule density $l_i\ll 1$, we have $x_i=1$ in the few territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. In our more general approximation, territories with a single propagule from the focal genotype are handled separately. In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion''). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq. \eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$). The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Thus, Eq. \eqref{eq:master} generalizes the classic lottery model to account for arbitrary propagule densities for each genotype.

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$). Two genotypes are present, one of which has a $c$-advantage and is at low frequency. The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b). On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d). Fig. 3 also shows the breakdown of the classic lottery model at low densities (low $l_1$ and $l_2$).

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq. \eqref{eq:master}. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present. (a) and (b) show the low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$). Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in (a) and (c) show the corresponding classic lottery model predictions.} 
\end{figure}

@@

Stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
